Prefetching SPARQL query cacher

### 2016-07-22 dissertation/cache-proxy.tex
Helen's corrections

The system is in the convergence of the directions described in this
dissertation: It should relieve the remote endpoint of some of the
burden to evaluate the query and it should add to the robustness of
the open Web SPARQL infrastructure. It uses hypermedia to answer
individual triple patterns, based on Triple Pattern Fragments (TPF)
\cite{ldf1}. The system does not support named graphs (i.e. only
triple stores), but other than that, supports SPARQL 1.1 in its
entirety. This was made possible to develop quickly by the efforts put
into the query planning in the Attean framework described in
Section~\ref{sec:conpush}. Caching based on RFC7234 \cite{rfc7234} was
shown to be of possible use in the survey I conducted. Finally, it was
planned to be evaluated using Design of Experiments.

@@

\draw [->] (0.2,4) --(0.2,0);
\draw [<-] (0.4,4) --(0.4,0);
\node at (0.3,4.5) {\circled{1}};
\draw [->] (0.9,2) --(0.9,0);
\draw [<-] (1.1,2) --(1.1,0);
\node at (1.0,4.5) {\circled{2}};
\draw [->] (1.6,4) --(1.6,2);
\draw [dashed, ->] (1.6,2) --(1.6,0);
\draw [dashed, <-] (1.8,2) --(1.8,0);
\draw [->] (1.9,2) --(1.9,4);
\node at (1.7,4.5) {\circled{3}};
\draw [->] (2.4,2) --(2.4,0);
\draw [<-] (2.6,2) --(2.6,0);
\node at (2.5,4.5) {\circled{4}};
\draw [->] (3,4) --(3,2);
\draw [->] (3.2,2) --(3.2,4);
\node at (3.1,4.5) {\circled{5}};
\end{tikzpicture}
\caption{The possible positions of a cache. A cache may reside at both
  the server side (as a database cache) or client side, or any of the
  intermediate proxies. The arrows illustrate requests (when pointing
  down) and responses (when pointing up) in a deployment scenario for
  when a the query cache considered in this paper is deployed on a
  caching proxy in the Internet infrastructure, see the text for
  details.}\label{fig:messaging}
\end{center}
\end{figure}

@@

Figure~\ref{fig:messaging} illustrates the case where HTTP messages
are being passed, where the cache is assumed to reside on a
intermediate caching proxy in the Internet, i.e. a CDN. The figure may
be viewed as having time on the $x$-axis, with the arrows pointing
down being requests and the arrows point up being responses. In this
example, the client first makes a request (request \circled{1}), which
is passed through the proxy because it has not yet cached any
responses. The response (response \circled{1}) is then also
transmitted directly from the SPARQL Endpoint at the database back to
the client. The entire response may be cached at the proxy in case it
can be reused in its entirety. However, the main point is that the
request will be analysed by the proxy in parallel to being sent to the
server, and another request \circled{2} for a triple pattern fragment
that the analyser thinks will enable the proxy to assist the server
the best for future requests, is sent. This analysis is discussed in
Section~\ref{sec:analpre}. The response \circled{2} is not sent back
to the client, but will enter the cache. Lets say that the analyser
was successful and that the cache can be used to answer request
\circled{3}. In that case, the proxy will send a rewritten query for
the rest of the data it needs to answer the query to the origin
server, indicated by the dashed arrows. This may be a SPARQL query, a
TPF request or a combination thereof, depending on what the planner
finds least expensive. The proxy will then evaluate the entire query
and send the final result back as response \circled{3}. In addition,
the analyser may have identified another Triple Pattern Fragment that
it takes as likely useful for future requests, and sends a request and
caches the response \circled{4}. At some point, it may be able to
answer the entire query using only cached results, as illustrated by
request and response \circled{5}.

@@

The implementation consists of several modules that have been
published to the Comprehensive Perl Archive Network as Free Software,
and most of which have become a part the Free Software ecosystem. To
install the system on the top of a Debian system with only essential
packages requires 146 Debian packages. Running the test suite requires
even more. As such, it builds on the code of hundreds of authors, too
long to enumerate. I shall constrain myself to list the modules that
have code that has been motivated from this study. Modules that have
other primary authors than myself have been started prior to this
present project, but may have substantial contributions from this
project. The Attean framework was started as part of the effort in
Section~\ref{sec:conpush}, but even though it has been developed
further with the requirements that arose from this project and has
contributions from me, the vast majority of the code has been written
by Gregory Todd Williams. The main work in this project has gone into
the module \pmodule{AtteanX::Query::Cache}, where I have authored the
vast majority of the code. The relative difference in authorship can
be further examined by checking the linked Github repositories, or by
using git2prov\footnote{See \url{http://git2prov.org/}} to create RDF
data of the commit history. Table~\ref{tab:modules} shows the modules
that have been influenced or written as part of this project.

@@

The system also depends on modules I have written that is not part of
the project: \pmodule{RDF::LinkedData}, \pmodule{RDF::Generator::Void}, \pmodule{Test::RDF},
\pmodule{URI::NamespaceMap} and \pmodule{RDF::NS::Curated}.

@@

Traits are groups of methods that serve as a primitive unit of code
reuse. Traits are not constructed to objects themselves and do not
use inheritance for composition, instead a class is composed by a set of
traits and possibly the class' own methods, and then instantiated. A
trait can also require a set of methods or attributes, and as such, is similar to
interfaces, but since its focus is code reuse, it can also provide a
default implementation.

@@

We have found it convenient to use classical inheritance in
combination with traits-based composition, but only when the base
class provide only fundamental functionality that is very likely to be
common to all implementations. For an example of this, see
Section~\ref{sec:implqueryplan}.

@@

The \pmodule{AtteanX::Store::SPARQL} module is a partial SPARQL
Protocol client that implements Attean APIs. The store implementation
itself composes the TripleStore API and implements methods to retrieve
the results of a single triple pattern and exact cardinality for a
given triple pattern, by using an aggregate query. In addition, it can
generate plans for Basic Graph Pattern algebra objects that have more
than one triple pattern. If that is the case, it will return an
instance of the class \pmodule{AtteanX::Plan::SPARQLBGP}, which is
also defined by the module. If it is not the case, the module also has
an implementation of the \pcode{access\_plan} method in a
\pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} role, that can be
composed by a query planner to provide a single triple pattern
\pmodule{AtteanX::Plan::SPARQLBGP} object. Finally, it also provides a
model implementation that composes the Model and CostPlanner
APIs. This contains a \pcode{cost\_for\_plan} implementation that will
provide a cost for \pmodule{AtteanX::Plan::SPARQLBGP} objects that are
proportional to the number of triple patterns in the plan, but
penalises plans that does not have triple patterns that are connected
through a variable (i.e. will cause a Cartesian join) with a factor
10. See Section~\ref{sec:costheuristics} for details on the cost
heuristics.

@@

The cache itself is assumed to only cache triples where at least one
term is bound. I have used the Redis data structure store for the
cache as well. This choice is somewhat arbitrary, and has not proved
very successful. The cache will store the results as an array if two
RDF terms were bound in the triple pattern, or a hash if only one term
was bound. The strings put into the cache are serialized N-Triples
strings.

@@

\pmodule{Attean::QueryPlanner} has only minimal functionality,
everything else is provided by the roles. Conventional single
inheritance is constrained to tree structure, and would result in a
very complex system. Another option is to use the factory method
pattern, where a factory object would need to be initialised to return
a suitable object for each axis of customisation.

@@

However, since the combination of remotely executed SPARQL with a
local cache and TPF is rather complex, the cost
heuristics are fairly elaborate. Moreover, since the objective of the
study is to take load off of the remote endpoint, the heuristics is
intended to make the proxy work harder than the remote endpoint.

@@

Next, we detail the cost of a join where one of more children in the
tree is a remote SPARQL plan. It is usually not wanted to execute
several remote queries, it is assumed that the number of HTTP requests
should be small, and that plans where the joins have been rotated and
coalesced by \pmodule{AtteanX::API::JoinRotatingPlanner} should be
less expensive. Therefore, such plans should be penalised, but by how
much depends on the type plans. Conventionally, for a nested loop join
plan, we multiply the cost of the subplans, increment the resulting
cost by one if the right hand plan is more expensive than the left
hand plan, and multiply by a factor of 10 if there are no shared
variables, to mildly penalise Cartesian joins. If the plan is a hash
join the difference is that, we add rather than multiply the cost of
the of the subplans, and multiply by a factor of 100 rather than 10 so
that if it is necessary to evaluate a Cartesian join, it is more
likely done using a nested loop join. Then, the resulting cost is
multiplied by the number of remote SPARQL Basic Graph Patterns in the
tree and the factor 1.2, and floored to the nearest integer.

@@

The system was developed in a practically oriented test-driven manner,
with emphasis on testing cases that I assumed \textit{a priori} to be
difficult. The interested reader is referred to the published code,
given in Section~\ref{sec:impl}. Some of the tests consist of a number
of queries, and some of the tests concerns the problem I found in a
preliminary study, that the cached triples would often break
subsequent queries to cause Cartesian joins. Much of the above cost
model is designed to prevent this from having a detrimental impact.

While I am quite confident that the code is producing query plans as
intended and that analysis and prefetching is able to retrieve and
cache results, an elaborate evaluation would be needed to further
explore whether the system actually helps remote endpoints and is able
to serve clients with adequate performance. Unfortunately, the
evaluation showed that the latter is not the case with the current
implementation, and the practical constraints of the project then
prevented further investigation. Thus, this section consists of an
account of the evaluation that were performed and the methodology I
planned to employ.

@@

In more detail, query logs can be obtained by from
e.g. USEWOD~\cite{eps385344} or the Linked SPARQL Queries
Dataset \cite{Saleem2015}. An extraction of one of those datasets
could be performed, to create a representative progression of
queries. The data normally residing on a remote endpoint
(i.e. ``server'') would then be loaded into a single database system
under my control. A proxy system would be set up using our
system. Finally, a client would submit the queries to the server, via
the proxy.

On the server, some metric of system load would recorded. In
statistical experimentation parlance, this is known as the ``response
variable''. I had not entirely decided which metric to use, but
\cite{Ferrari:1987:EIL:647412.725013} indicated the load average may
be a good choice. It could be trivially recorded by polling the
virtual file \texttt{/proc/loadavg} on the host system. Possibly a
correction for any extra time spent by the proxy could be applied.

As for the overall design of the experiment, the reader is referred to
the paper discussed in Section~\ref{sec:condoe} for an introduction to
the type of evaluation intended to be used. As noted, a $2^n$
factorial experiment is the simplest experiment to set up, and I would
therefore go to great lengths to ensure that such an experiment could
be used.

@@

While this performed better in most cases, the test suite contains a
query where it does not:
\begin{verbatim}
?s <http://example.org/m/p> "1" ;
   <http://example.org/m/p> ?o ;
   <http://example.org/m/q> _:xyz ;
   <http://example.org/m/q> <http://example.org/m/a> .
?a <http://example.org/m/b> <http://example.org/m/c> .
\end{verbatim}
Note that the above Basic Graph Pattern contains a triple pattern that
does not share a variable (\sparql{?a}) with any other triple pattern,
thus creating a problematic Cartesian join. Additionally, in this
test, the results of the two first triple patterns have been
cached. \textit{A priori}, I have assumed that the best plan is a plan
where the cache is used for these two triple patterns, the last triple
pattern is answered by a LDF server and the last two
triple patterns are coalesced into a new SPARQL query, and evaluated
remotely. This is also coded as a pass criterion for the test. With
the simple query planner, this test passes, but not with the IDP
planner, as it results in a plan where the two triple patterns that
should run is a single SPARQL query is split into two. This resulted
in an interesting observation:

As noted in Section~\ref{sec:implqueryplan}, we rely on rotating joins
to gather triple pattern plans into a larger structure that can be
evaluated together, like a remote SPARQL BGP. On careful inspection, I
found that the join rotation would have had to descend into
great-grandchildren of the top join plan. However, with the IDP
planner, plans are aggressively pruned, and therefore, this
constellation is not considered. This led us (with Gregory Todd
Williams) to conjecture that this weakness is inherit to IDP planning
(and possibly a general problem with any early-pruning planner) 
when using join rotation. We did not attempt to examine this any
further, however.

To remedy this problem, we considered creating a custom query planner
for Basic Graph Patterns, but I did not implement it. The outline
would be as follows:
\begin{enumerate}
\item Find triple patterns that are connected with a common variable,
  and group them to ``components''.
\item If a triple pattern have a cached result, create a plan that
  accesses the cache. 
\item If the triple pattern in the previous step caused a Cartesian
  join to be created within a component, create an equivalent
  additional plan for a SPARQL remote evaluation that does not access
  the local cache.
\item Any remaining components consisting of a single triple pattern
  result in LDF plans.
\item Any remaining components are turned into SPARQL remote BGP
  plans.
\end{enumerate}

@@

One may raise the issue that it was a bad idea to begin with to use a
scripting language like Perl to write a query engine with performance
requirements. Indeed, that may be the case. However, generally
accepted practice is to first find the actual performance issues and
then optimise using a bridge to first and foremost the C family of
languages as needed. Importantly, with the width of the scope, I only
had some months to implement the query engine. This is not a software
engineering project, but given past industrial experience, it seems
unlikely that I could have achieved it within the allowed time frame
using older paradigms, such as the factory method pattern.

@@

However, since computing the join itself is too slow, it is not
necessarily a solution worth pursuing. The main motivation behind
caching hashes and arrays in Redis was to save space; since at least
one term would be bound in any cached pattern, it shouldn't be
necessary to insert more data into the cache. Since the performance of
doing the join is poor, this may not be justified. Instead, it may be
a better idea to use a conventional quad store for the cache. In that
case, the planner would require a simple modification so that triple
patterns are gathered into locally evaluated BGPs and remotely
evaluated BGPs as well as LDFs. To take full advantage of this
approach, SPARQL 1.1 \sparql{VALUES} would also need to be supported,
so that the local quad store could perform all join operations. To
this end, I wrote a rudimentary Virtuoso store driver for Attean, but
at that point, the project was overdue and needed to be concluded.

It seems likely that the project, with more time, could be brought to
the point where a the evaluation in Section~\ref{sec:planned} could be
applied, and new insight could be gained.

>>>>

The system is in the convergence of the directions described in this
thesis: It should relieve the remote endpoint of some of the
burden to evaluate the query and it should add to the robustness of
the open Web SPARQL infrastructure. It uses hypermedia to answer
individual triple patterns, based on Triple Pattern Fragments (TPF)
\cite{ldf1}. The system does not support named graphs (i.e. only
triple stores), but other than that, it supports SPARQL 1.1 in its
entirety. It was possible to develop this quickly by the efforts put
into the query planning in the Attean framework described in
Section~\ref{sec:conpush}. Caching based on RFC7234 \cite{rfc7234} was
shown to be of possible use in the survey I conducted. Finally, it was
planned to be evaluated using Design of Experiments.

@@

\draw [->] (0.2,4) --(0.2,0);
\draw [<-] (0.4,4) --(0.4,0);
\node at (0.3,4.5) {\circled{1}};
\draw [->] (0.9,2) --(0.9,0);
\draw [<-] (1.1,2) --(1.1,0);
\node at (1.0,4.5) {\circled{2}};
\draw [->] (1.6,4) --(1.6,2);
\draw [dashed, ->] (1.6,2) --(1.6,0);
\draw [dashed, <-] (1.8,2) --(1.8,0);
\draw [->] (1.9,2) --(1.9,4);
\node at (1.7,4.5) {\circled{3}};
\draw [->] (2.4,2) --(2.4,0);
\draw [<-] (2.6,2) --(2.6,0);
\node at (2.5,4.5) {\circled{4}};
\draw [->] (3,4) --(3,2);
\draw [->] (3.2,2) --(3.2,4);
\node at (3.1,4.5) {\circled{5}};
\end{tikzpicture}
\caption{The possible positions of a cache. A cache may reside at both
  the server side (as a database cache) or client side, or any of the
  intermediate proxies. The arrows illustrate requests (when pointing
  down) and responses (when pointing up) in a deployment scenario for
  when the query cache considered in this paper is deployed on a
  caching proxy in the Internet infrastructure (see the text for
  details).}\label{fig:messaging}
\end{center}
\end{figure}

@@

Figure~\ref{fig:messaging} illustrates the case where HTTP messages
are being passed, where the cache is assumed to reside on a
intermediate caching proxy in the Internet, i.e. a CDN. The figure may
be viewed as having time on the $x$-axis, with the arrows pointing
down being requests and the arrows point up being responses. In this
example, the client first makes a request (request \circled{1}), which
is passed through the proxy because it has not yet cached any
responses. The response (response \circled{1}) is then also
transmitted directly from the SPARQL Endpoint at the database back to
the client. The entire response may be cached on the proxy in case it
can be reused in its entirety. However, the main point is that the
request will be analysed by the proxy in parallel to being sent to the
server, and another request \circled{2} for a triple pattern fragment
that the analyser thinks will enable the proxy to assist the server
the best for future requests, is sent. This analysis is discussed in
Section~\ref{sec:analpre}. The response \circled{2} is not sent back
to the client, but will enter the cache. Lets say that the analyser
was successful and that the cache can be used to answer request
\circled{3}. In that case, the proxy will send a rewritten query for
the rest of the data it needs to be able to answer the query to the origin
server, indicated by the dashed arrows. This may be a SPARQL query, a
TPF request or a combination thereof, depending on what the planner
finds least expensive. The proxy will then evaluate the entire query
and send the final result back as response \circled{3}. In addition,
the analyser may have identified another Triple Pattern Fragment that
it takes as likely to be useful for future requests, and sends a request and
caches the response \circled{4}. At some point, it may be able to
answer the entire query using only cached results, as illustrated by
request and response \circled{5}.

@@

The implementation consists of several modules that have been
published to the Comprehensive Perl Archive Network as Free Software,
and most of which have become a part the Free Software ecosystem. To
install the system on the top of a Debian system with only essential
packages requires 146 Debian packages. Running the test suite requires
even more. As such, it builds on the code of hundreds of authors, too
many to enumerate. I shall constrain myself to list the modules that
have code that has been motivated by this study. Modules that have
other primary authors than myself have been started prior to this
present project, but may have substantial contributions from this
project. The Attean framework was started as part of the effort in
Section~\ref{sec:conpush}, but even though it has been developed
further with the requirements that arose from this project and has
contributions from me, the vast majority of the code has been written
by Gregory Todd Williams. The main work in this project has gone into
the module \pmodule{AtteanX::Query::Cache}, where I have authored the
vast majority of the code. The relative difference in authorship can
be further examined by checking the linked Github repositories, or by
using git2prov\footnote{See \url{http://git2prov.org/}} to create RDF
data of the commit history. Table~\ref{tab:modules} shows the modules
that have been influenced or written as part of this project.

@@

The system also depends on modules I have written that are not part of
the project: \pmodule{RDF::LinkedData}, \pmodule{RDF::Generator::Void}, \pmodule{Test::RDF},
\pmodule{URI::NamespaceMap} and \pmodule{RDF::NS::Curated}.

@@

Traits are groups of methods that serve as a primitive unit of code
reuse. Traits are not constructed into objects themselves and do not
use inheritance for composition, instead a class is composed by a set
of traits and possibly the class' own methods, and then
instantiated. A trait can also require a set of methods or attributes,
and as such, is similar to an interface, but since its focus is code
reuse, it can also provide a default implementation.

@@

We have found it convenient to use classical inheritance in
combination with traits-based composition, but only when the base
class provide just fundamental functionality that is very likely to be
common to all implementations. For an example of this, see
Section~\ref{sec:implqueryplan}.

@@

The \pmodule{AtteanX::Store::SPARQL} module is a partial SPARQL
Protocol client that implements Attean APIs. The store implementation
itself composes the TripleStore API and implements methods to retrieve
the results of a single triple pattern and exact cardinality for a
given triple pattern, by using an aggregate query. In addition, it can
generate plans for Basic Graph Pattern algebra objects that have more
than one triple pattern. If that is the case, it will return an
instance of the class \pmodule{AtteanX::Plan::SPARQLBGP}, which is
also defined by the module. If it is not the case, the module also has
an implementation of the \pcode{access\_plan} method in a
\pmodule{AtteanX::Query::AccessPlan::SingleQuadBGP} role, that can be
composed by a query planner to provide a single triple pattern
\pmodule{AtteanX::Plan::SPARQLBGP} object. Finally, it also provides a
model implementation that composes the Model and CostPlanner
APIs. This contains a \pcode{cost\_for\_plan} implementation that will
provide a cost for \pmodule{AtteanX::Plan::SPARQLBGP} objects that are
proportional to the number of triple patterns in the plan, but
penalises plans that do not have triple patterns that are connected
through a variable (i.e. will cause a Cartesian join) with a factor
10. See Section~\ref{sec:costheuristics} for details on the cost
heuristics.

@@

The cache itself is assumed to only cache triples where at least one
term is bound. I have used the Redis data structure store for the
cache as well. This choice is somewhat arbitrary, and has not proved
very successful. The cache will store the results as an array if two
RDF terms are bound in the triple pattern, or a hash if only one term
is bound. The strings put into the cache are serialized N-Triples
strings.

@@

\pmodule{Attean::QueryPlanner} has only minimal functionality,
everything else is provided by the roles. Conventional single
inheritance is constrained to a tree structure, and would result in a
very complex system. Another option is to use the factory method
pattern, where a factory object would need to be initialised to return
a suitable object for each axis of customisation.

@@

However, since the combination of remotely executed SPARQL with a
local cache and TPF is rather complex, the cost
heuristics are fairly elaborate. Moreover, since the objective of the
study is to take load off of the remote endpoint, the heuristics are
intended to make the proxy work harder than the remote endpoint.

@@

Next, we detail the cost of a join where one or more children in the
tree is a remote SPARQL plan. It is usually not wanted to execute
several remote queries, it is assumed that the number of HTTP requests
should be small, and that plans where the joins have been rotated and
coalesced by \pmodule{AtteanX::API::JoinRotatingPlanner} should be
less expensive. Therefore, such plans should be penalised, but by how
much depends on the type plans. Conventionally, for a nested loop join
plan, we multiply the cost of the subplans, increment the resulting
cost by one if the right hand plan is more expensive than the left
hand plan, and multiply by a factor of 10 if there are no shared
variables, to mildly penalise Cartesian joins. If the plan is a hash
join the difference is that, we add rather than multiply the cost of
the of the subplans, and multiply by a factor of 100 rather than 10 so
that if it is necessary to evaluate a Cartesian join, it is more
likely done using a nested loop join. Then, the resulting cost is
multiplied by the number of remote SPARQL Basic Graph Patterns in the
tree and the factor 1.2, and floored to the nearest integer.

@@

The system was developed in a practically oriented test-driven manner,
with emphasis on testing cases that I assumed \textit{a priori} to be
difficult. The interested reader may like to refer to the published code,
given in Section~\ref{sec:impl}. Some of the tests consist of a number
of queries, and some of the tests concern the problem I found in a
preliminary study, that the cached triples would often break
subsequent queries to cause Cartesian joins. Much of the above cost
model is designed to prevent this from having a detrimental impact.

While I am quite confident that the code is producing query plans as
intended and that analysis and prefetching is able to retrieve and
cache results, an elaborate evaluation would be needed to explore
further whether the system actually helps remote endpoints and is able
to serve clients with adequate performance. Unfortunately, the
evaluation showed that the latter is not the case with the current
implementation, and the practical constraints of the project then
prevented further investigation. Thus, this section consists of an
account of the evaluation that was performed and the methodology I
planned to employ.

@@

In more detail, query logs can be obtained from
for example USEWOD~\cite{eps385344} or the Linked SPARQL Queries
Dataset \cite{Saleem2015}. An extraction of one of those datasets
could be performed, to create a representative progression of
queries. The data normally residing on a remote endpoint
(i.e. ``server'') would then be loaded into a single database system
under my control. A proxy system would be set up using our
system. Finally, a client would submit the queries to the server, via
the proxy.

On the server, some metric of system load would be recorded. In
statistical experimentation parlance, this is known as the ``response
variable''. I had not entirely decided which metric to use, but
\cite{Ferrari:1987:EIL:647412.725013} indicated the load average may
be a good choice. It could be trivially recorded by polling the
virtual file \texttt{/proc/loadavg} on the host system. Possibly a
correction for any extra time spent by the proxy could be applied.

As for the overall design of the experiment, the reader may refer to
the paper discussed in Section~\ref{sec:condoe} for an introduction to
the type of evaluation intended to be used. As noted, a $2^n$
factorial experiment is the simplest experiment to set up, and I would
therefore go to great lengths to ensure that such an experiment could
be used.

@@

While this performed better in most cases, the test suite contains a
query where it does not:
\begin{verbatim}
?s <http://example.org/m/p> "1" ;
   <http://example.org/m/p> ?o ;
   <http://example.org/m/q> _:xyz ;
   <http://example.org/m/q> <http://example.org/m/a> .
?a <http://example.org/m/b> <http://example.org/m/c> .
\end{verbatim}
Note that the above Basic Graph Pattern contains a triple pattern that
does not share a variable (\sparql{?a}) with any other triple pattern,
thus creating a problematic Cartesian join. Additionally, in this
test, the results of the two first triple patterns have been
cached. \textit{A priori}, I have assumed that the best plan is a plan
where the cache is used for these two triple patterns, the last triple
pattern is answered by a LDF server and the last two
triple patterns are coalesced into a new SPARQL query, and evaluated
remotely. This is also coded as a pass criterion for the test. With
the simple query planner, this test passes, but not with the IDP
planner, as it results in a plan where the two triple patterns that
should run is a single SPARQL query are split into two. This resulted
in an interesting observation:

As noted in Section~\ref{sec:implqueryplan}, we rely on rotating joins
to gather triple pattern plans into a larger structure that can be
evaluated together, like a remote SPARQL BGP. On careful inspection, I
found that the join rotation would have had to descend into
great-grandchildren of the top join plan. However, with the IDP
planner, plans are aggressively pruned, and therefore, this
constellation is not considered. This led us (with Gregory Todd
Williams) to conjecture that this weakness is inherit to IDP planning
(and possibly a general problem with any early-pruning planner) 
when using join rotation. However, we did not attempt to examine this any
further.

To remedy this problem, we considered creating a custom query planner
for Basic Graph Patterns, but this was not implemented. The outline
would have been as follows:
\begin{enumerate}
\item Find triple patterns that are connected with a common variable,
  and group them to ``components''.
\item If a triple pattern has a cached result, create a plan that
  accesses the cache. 
\item If the triple pattern in the previous step causes a Cartesian
  join to be created within a component, create an equivalent
  additional plan for a SPARQL remote evaluation that does not access
  the local cache.
\item Any remaining components consisting of a single triple pattern
  result in LDF plans.
\item Any remaining components are turned into SPARQL remote BGP
  plans.
\end{enumerate}

@@

The issue could be raised that it was a bad idea to begin with to use a
scripting language like Perl to write a query engine with performance
requirements. Indeed, that may be the case. However, generally
accepted practice is to first find the actual performance issues and
then optimise using a bridge to first and foremost the C family of
languages as needed. Importantly, with the width of the scope, I only
had some months to implement the query engine. This is not a software
engineering project, but given past industrial experience, it seems
unlikely that I could have achieved it within the allowed time frame
using older paradigms, such as the factory method pattern.

@@

However, since computing the join itself is too slow, it is not
necessarily a solution worth pursuing. The main motivation behind
caching hashes and arrays in Redis was to save space; since at least
one term would be bound in any cached pattern, it shouldn't be
necessary to insert more data into the cache. Since the performance of
doing the join is poor, this may not be justified. Instead, it may be
a better idea to use a conventional quad store for the cache. In that
case, the planner would require a simple modification so that triple
patterns are gathered into locally evaluated BGPs and remotely
evaluated BGPs as well as LDFs. To take full advantage of this
approach, SPARQL 1.1 \sparql{VALUES} would also need to be supported,
so that the local quad store could perform all join operations. With
this in mind, I wrote a rudimentary Virtuoso store driver for Attean, but
at that point, the project was overdue and needed to be concluded.

It seems likely that the project, with more time, could be brought to
the point where the evaluation in Section~\ref{sec:planned} could be
applied, and new insight could be gained.
